A general Nullstellensatz for generalized spaces

### 2019-03-09 paper-qcoh.tex
to do
typo

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
any nonzero element is invertible.
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings, suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

@@

Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
hypothesis. This principle holds for instance in the case that~$A$ is an
algebraically closed field and that~$g$ is the unit polynomial. It is also
true, without any restriction on~$g$, for the generic ring.

We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
follows. \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
algebraic certificate we now have a logical certificate, a \emph{proof}
of~$\sigma$. However, this generalized statement is typically false, even for
the generic model~$U_\TT$: The statement
\begin{multline*}
  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
    proves~$\sigma$}\urcorner\qquad\qquad\qquad
\end{multline*}
does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
instance, if~$\TT$ is the theory of rings, then from the internal point of view
of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
More details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader.}

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory. We then have, subject to some qualifications made precise in
Section~\ref{sect:main}:

@@

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be of the form~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \in U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. We formulate a substitute as Theorem~XXX,
substantially broading the scope of the Nullstellensatz.

@@

A situation in which the sheaf assumption of Theorem~\ref{thm:nullstellensatz-horn}
is satisfied is when~$\TT$ is a Horn theory and the topology cutting
out~$\Set[\TT']$ is subcanonical. For instance, this is the case if~$\TT$ is
the theory of rings and~$\Set[\TT']$ is any of the well-known toposes in
algebraic geometry such as the big Zariski topos, the big étale topos or the
big fppf topos.

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$R$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.

We omit the required verifications and only remark that while the same
construction can be carried out if~$\TT$ is a general geometric theory, the
resulting object would in general not be a model of~$\TT$.
\end{proof}

@@

\begin{lemma}A~$\TT$-model is a finitely presented object of the category
of~$\TT$-models if and only if it is isomorphic to a model of the
form~$\TT\langle X | R \rangle$ where~$X$ is Bishop-finite and~$R$ is
Kuratowski-finite.
\end{lemma}

@@

\begin{lemma}Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
with~$yA$, where~$y : \Mod{\TT} \to \Set[\TT]$ is the Yoneda embedding.
\end{lemma}

@@

@@

>>>>

From the point of view of the topos, the ring object~$\affl$ is not only a
local ring, but even a field in the sense that
any nonzero element is invertible.
As this condition is of nongeometric form, it is not inherited by arbitrary
local rings, which are indeed typically not fields. However, any intuitionistic
consequence of this condition which is of geometric form is inherited
by any local ring in any topos. Hence we may, when verifying a general fact
about local rings which is expressible as a geometric sequent,
suppose without loss of generality that the given ring is a
field. This observation is due to Kock~\cite{kock:univ-proj-geometry}, who exploited it to
develop projective geometry over local rings, and was further used by Reyes to
prove a Jacobian criterion for étale morphisms~\cite{reyes:cramer}.

@@

Let~$A$ be a ring. A simple version of the classical Nullstellensatz states:
\emph{For any polynomials~$f$ and~$g$ over~$A$, if any zero of~$f$ is also a
zero of~$g$, then there is a polynomial~$h$ such that~$g = hf$.} The
polynomial~$h$ can be regarded as an ``algebraic certificate'' of the
hypothesis. This principle holds for instance in the case that~$A$ is an
algebraically closed field and that~$g$ is the unit polynomial. We will see
below that it is also
true, without any restriction on~$g$, for the generic ring.

We could try to generalize the Nullstellensatz to arbitrary geometric theories~$\TT$ as
follows: \emph{For any geometric sequent~$\sigma$, if~$\sigma$ holds for a
given~$\TT$-model~$M$ then~$\sigma$ is provable modulo~$\TT$.} In place of the
algebraic certificate we now have a logical certificate, a \emph{proof}
of~$\sigma$. However, this generalized statement is typically false, even for
the generic model~$U_\TT$: The statement
\begin{multline*}
  \qquad\qquad\qquad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
    \text{if~$\sigma$ holds for~$U_\TT$ then~$\ul{\TT}$
    proves~$\sigma$}\urcorner\qquad\qquad\qquad
\end{multline*}
does not hold.\footnote{Here~$\ul{\TT}$ is the internal geometric theory induced
by~$\TT$, obtained by pulling back the set of sorts, the set of function
symbols and so on along the geometric morphism~$\Set[\TT] \to \Set$. For
instance, if~$\TT$ is the theory of rings, then from the internal point of view
of~$\Set[\TT]$ the theory~$\ul{\TT}$ will again be the theory of rings.
More details will be given in Section~\ref{sect:review}. The corner quotes indicate
that for sake of readability, the translation into formal language is to be
carried out by the reader. \par The displayed statement is much stronger than
the statement that for any geometric sequent~$\sigma$, if $\Set[\TT] \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$\TT$ proves~$\sigma$. This latter statement,
where the universal quantifier and the ``if \ldots then'' have been pulled out,
is true.}

@@

Intuitively, the problem is that while the meaning of~$\speak{$\ul{\TT}$
proves~$\sigma$}$ is fixed, the meaning of~$\speak{$\sigma$ holds for~$U_\TT$}$ can vary
with the slice. This problem can be solved by passing from~$\ul{\TT}$ to a
varying theory, the internal theory~$\ul{\TT}/U_\TT$ defined in
Section~\ref{sect:main}. If~$\TT$ is the theory of rings, then~$\ul{\TT}/U_\TT$
is the~$\Set[\TT]$-theory of~$U_\TT$-algebras.
Unlike~$\ul{\TT}$, this theory is not the pullback of an external geometric
theory. We then have, subject to some qualifications made precise in
Section~\ref{sect:main}, the following general Nullstellensatz:

@@

To illustrate Theorem~\ref{thm:nullstellensatz0}, let~$\TT$ be the theory of
rings and let~$\sigma$ be the sequent~$(f(x) = 0 \seq{x} g(x) = 0)$ for some
polynomials~$f$ and~$g$. To say that~$\sigma$ holds for~$U_\TT$ amounts to
saying that any zero~$x \? U_\TT$ of~$f$ is also a zero of~$g$, and to say
that~$\ul{\TT}/U_\TT$ proves~$\sigma$ amounts to saying that
in~$U_\TT[X]/(f(X))$, the free~$U_\TT$-algebra on one generator~$X$ subject to
the relation~$f(X) = 0$, the relation~$g([X]) = 0$ holds. Hence we obtain
\[ \Set[\TT] \models
  \forall f,g \? U_\TT[X]\_ \bigl(
    (\forall x \? U_\TT\_ f(x) = 0 \Rightarrow g(x) = 0) \Longleftrightarrow
      \exists h \in U_\TT[X]\_ g = hf\bigr). \]

The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. We formulate a substitute as Theorem~XXX,
substantially broadening the scope of the Nullstellensatz.

@@

A situation in which the sheaf assumption of Theorem~\ref{thm:nullstellensatz-horn}
is satisfied is when~$\TT$ is a Horn theory and the topology cutting
out~$\Set[\TT']$ is subcanonical. For instance, this is the case if~$\TT$ is
the theory of rings and~$\Set[\TT']$ is one of several well-known toposes in
algebraic geometry such as the big Zariski topos, the big étale topos or the
big fppf topos.

@@

\begin{proof}The desired model can be constructed as a term algebra. As a set,
it consists of the terms (in the empty context) of the signature~$\Sigma + X$
modulo the equivalence relation identifying two terms if and only if~$\TT + R$
proves them to be equal. The function symbols~$f$ of~$\Sigma$ are interpreted
by declaring~$\brak{f}([t_1],\ldots,[t_n]) = [f(t_1,\ldots,t_n)]$ and the
relation symbols~$S$ are interpreted by declaring~$([t_1],\ldots,[t_n]) \in
\brak{S} \Leftrightarrow (\TT + R \vdash S(t_1,\ldots,t_n))$.

We omit the required verifications and only remark that while the same
construction could be carried out if~$\TT$ is a general geometric theory, the
resulting object would in general not be a model of~$\TT$.
\end{proof}

@@

\begin{lemma}A~$\TT$-model is finitely presentable as an object of the category
of~$\TT$-models if and only if it is isomorphic to a model of the
form~$\TT\langle X | R \rangle$ where~$X$ is Bishop-finite and~$R$ is
Kuratowski-finite.
\end{lemma}

@@

\begin{lemma}Let~$A$ be a~$\TT$-model. Then~$\Spec(A^\sim)$ coincides
with~$yA$, the functor~$\Hom_{\Mod{\TT}}(A, \cdot)$.
\end{lemma}

@@

* Discuss related work, especially: Olivia 6.1.3, ...

@@

* Anmerken, dass wir in konstruktiver Metalogik arbeiten.
